# Effective Visualizations

Now that you know how to create graphics and visualizations in R, you are armed with powerful tools for scientific computing and analysis. With this power also comes great responsibility. Effective visualizations is an incredibly important aspect of scientific research and communication. There have been several books (see references) written about these principles. In class today we will be going through several case-studies trying to develop some expertise into making effective visualizations. 

## Worksheet

**The worksheet questions for today are embedded into the class notes.**

You can download this Rmd file [here](https://github.com/STAT545-UBC/Classroom/raw/master/cm013.Rmd)

Note, there will be very little coding in-class today, but I've given you plenty of exercises in the form of a supplemental worksheet (linked at the bottom of this page) to practice with after class is over.

## Resources

1. [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/introduction.html) by Claus Wilke.

1. [Visualization Analysis and Design](https://www-taylorfrancis-com.ezproxy.library.ubc.ca/books/9780429088902) by Tamara Munzner.

1. [STAT545.com - Effective Graphics](https://stat545.com/effective-graphs.html) by Jenny Bryan.

1. [ggplot2 book](https://ggplot2-book.org) by Hadley Wickam.

1. [Callingbull.org](https://callingbull.org/tools.html) by Carl T. Bergstrom and Jevin West.

## Part 1: Warm-up and pre-test [20 mins]

### Warmup:

Write some notes here about what "effective visualizations" means to you. Think of elements of good graphics and plots that you have seen - what makes them good or bad? Write 3-5 points.

1. Axes are effectively scaled, designed and labelled.
1. Colors can be used effectively
1. The data is clear to understand 
1. 
1.

### CQ01: Weekly hours for full-time employees

> Question: Evaluate the strength of the claim based on the data: "German workers are more motivated and work more hours than workers in other EU nations."
>
> Very strong, strong, weak, very week, do not know

- <<Weak. The distribution is quite small, only from 37 to 41, and does not show the standard deviation or size of the distribution within each country.>>

- Main takeaway: Summarize the main takeaway from this question/discussion here

### CQ02: Average Global Temperature by year

> Question: For the years this temperature data is displayed, is there an appreciable increase in temperature?
> 
> Yes

- <<Your answer here (and make sure to explain why you chose this answer)>>

- Main takeaway: Summarize the main takeaway from this question/discussion here

### CQ03: Gun deaths in Florida

> Question: Evaluate the strength of the claim based on the data: “Soon after this legislation was passed, gun deaths sharply declined."
>
> Very strong, strong, weak, very week, do not know

- Extremley weak. The graph is upside down, so the number of murders increased after "stand your ground", but it looks like  it went down because of the way the Y-axis is s designed. 
- Main takeaway: Summarize the main takeaway from this question/discussion here

## Part 2: Extracting insight from visualizations  [20 mins]

Great resource for selecting the right plot: https://www.data-to-viz.com/ ; encourage you all to consult it when choosing to visualize data.

### Case Study 1: Context matters

### Case Study 2: A case for pie charts

## Part 3: Principles of effective visualizations [20 mins]

We will be filling these principles in together as a class

1. The principle of proportional ink - the ink used should  be proportional to the value represented, and usually truncated axes are bad.
1. Maintain a high data to ink ratio.
1. Always maintain up to date axis labels and titles of the plots.
1. Choose the scaling systems carefully - linear, log, etc.
1. Choose graph type carefully 
1. Choose colors with accessiblity in mind- make it colorblind friendly and perceptually uniform. 

### Make a great plot worse

Instructions: Below is a code chunk that shows an effective visualization. First, copy this code chunk into a new cell. Then, modify it to purposely make this chart "bad" by breaking the principles of effective visualization above. Your final chart still needs to run/compile and it should still produce a plot. 

How many of the principles did you manage to break?

```{r}
ggplot(airquality, aes(`Month`, `Temp`, group = `Month`)) +
    geom_boxplot(outlier.shape = NA) +
    labs(x = "",
         y = "",
         title="Maximum temperature by month")+
    
    annotate("text", x = 4.08, y = 95,label="°F",size=8) +
    coord_cartesian(xlim = c(4.5, 9.5),
                      clip = 'off')+
    theme(panel.grid.minor = element_blank(),
          panel.background = element_blank(), 
          axis.line = element_line(colour = "gray"),
          panel.border = element_blank(),
          text = element_text(size=18)
          )
```



## Plotly demo [10 mins]

Did you know that you can make interactive graphs and plots in R using the plotly library? We will show you a demo of what plotly is and why it's useful, and then you can try converting a static ggplot graph into an interactive plotly graph.

This is a preview of what we'll be doing in STAT 547 - making dynamic and interactive dashboards using R!
```{r}
install.packages('plotly')
library(tidyverse)
library(gapminder)
library(plotly)
```


```{r}

p <- ggplot(gapminder, aes(x=gdpPercap, y=lifeExp, color = continent)) +
  geom_point()


#make interactive
p %>%
  ggplotly()

```

## Supplemental worksheet (Optional)

You are highly encouraged to the cm013 supplemental exercises worksheet. It is a great guide that will take you through Scales, Colours, and Themes in ggplot. There is also a short guided activity showing you how to make a ggplot interactive using plotly.

- [Supplemental Rmd file here](https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/tutorials/cm013-supplemental.Rmd)